---
title: "Getting Started with Cloud-Native Harmonized Landsat Sentinel-2 (HLS) Data in R"
output:
  html_document:
    df_print: paged
    fig_caption: yes
    theme: paper
    toc: yes
    toc_depth: 2
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
  word_document:
    toc: yes
    toc_depth: '2'
theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_knit$set(root.dir = dirname(rprojroot::find_rstudio_root_file()))
```

------------------------------------------------------------------------

**This tutorial demonstrates how to work with the HLS Landsat (HLSL30.002) and Sentinel-2 (HLSS30.002) data products in R.** 

The Harmonized Landsat Sentinel-2 [(HLS)](https://lpdaac.usgs.gov/data/get-started-data/collection-overview/missions/harmonized-landsat-sentinel-2-hls-overview)
project is a NASA initiative aiming to produce a consistent, harmonized
surface reflectance product from the Operational Land Imager (OLI) aboard 
the joint NASA/USGS Landsat 8 and Landsat 9 satellites and the Multi-Spectral 
Instrument (MSI) aboard Europe’s Copernicus Sentinel-2A and Sentinel-2B 
satellites. Using sets of algorithms, all the necessary radiometric, spectral,
geometric, and spatial corrections have been applied to make HLS into seamless
timeseries that are stackable and comparable.  

The dense timeseries of HLS data presents unprecedented opportunities to 
monitor and map land surface dynamics with exceptional spatial and temporal
detail. The enhanced resolution will provide value in numerous fields, 
including:  

- Land cover change detection
- Agricultural management and monitoring
- Disaster response and recovery
- Water resource management
- Vegetation phenology studies  

By leveraging the higher temporal resolution of this dataset, researchers and 
practitioners can gain a deeper understanding of complex environmental 
processes and make more informed decisions. 

NASA's Land Processes Distributed Active Archive Center (LP DAAC) archives 
and distributes HLS products in NASA's Earthdata cloud where each file is
stored as Cloud Optimized GeoTIFFs (COG). The primary objective of this 
tutorial is to show how to query and subset HLS data using the NASA CMR-STAC
application programming interface (API). Using cloud hosted publicly available 
data from NASA's Earthdata Cloud, you are not restricted by the need to 
download HLS data files for your research needs anymore. The CMR-STAC API 
allows you to subset your desired data collection by region, time, and band 
while Earthdata Cloud allows you to download or stream the desired data to your 
workspace.  

---

## Use Case Example

In this tutorial, we demonstrate how to process and analyze NDVI time series
data derived from HLS data over a region of interest (ROI). Using a case study,
we cover the following steps:

- **Calculate NDVI**: Learn how to compute the Normalized Difference Vegetation
Index (NDVI).
- **Quality Masking**: Apply quality masks to filter out low-quality data.
- **Visualization**: Visualize the NDVI data for better understanding and
interpretation.
- **Statistical Analysis**: Calculate and export statistics specific to your
ROI without downloading the source data files.  

We use multiple agricultural fields in California’s Central Valley as an 
example to illustrate how to interact with HLS data effectively.

### Products Used:	
	
**1. Daily 30 meter (m) global HLS Sentinel-2 Multi-spectral Instrument Surface Reflectance - [HLSS30.002](https://doi.org/10.5067/HLS/HLSS30.002)**    
   **Science Dataset (SDS) layers:**    
    - B8A (NIR Narrow)    
    - B04 (Red)    
    - Fmask (Quality)  

**2. Daily 30 meter (m) global HLS Landsat-8 and 9 OLI Surface Reflectance - [HLSL30.002](https://doi.org/10.5067/HLS/HLSL30.002)**  
  **Science Dataset (SDS) layers:**  
    - B05 (NIR)  
    - B04 (Red)  
    - Fmask (Quality)  

## Topics Covered in This Tutorial

1.  **Getting Started**\
    1a. Load Required Libraries\
    1b. Set Up the Working Directory\
2.  **CMR-STAC API: Searching for Items**\
    2a. Collection Query\
    2b. Spatial Query Search Parameter\
    2c. Temporal Query Search Parameter\
    2d. Submit a Query for Our Search Criteria\
3.  **Accessing and Interacting with HLS Data**\
    3a. Subset by Band\
    3b. Subset HLS COGs Spatially and Stack HLS Data Layers\
4.  **Processing HLS Data**\
    4a. Calculate NDVI\
    4b. Apply Quality Filtering\
    4c. Visualize Stacked Time Series\
    4d. Export Statistics\
    5.  **Export Output to GeoTIFF**  

## Prerequisites:

-   This tutorial can be executed using R with RStudio or in Visual Studio Code
(VS Code).
-   Tested on Windows using R Version 4.0.5 and RStudio version 1.2.5033.
-   A [NASA Earthdata Login](https://urs.earthdata.nasa.gov/) account is 
required to access the data used in this Tutorial. You can create an account
[here](https://urs.earthdata.nasa.gov/users/new).  

## Procedures:

### Getting Started:

  - [Clone](https://github.com/nasa/HLS-Data-Resources.git) or 
    [download](https://github.com/nasa/HLS-Data-Resources/archive/refs/heads/main.zip)
    the HLS Data Resources Repository. 

  - When you open this Rmarkdown notebook in RStudio, you can click the little 
    green "Play" button in each grey code chunk to execute the code. The result
    can be printed either in the R Console or inline in the RMarkdown notebook,
    depending on your RStudio preferences. 

### Environment Setup: 

#### 1. Check the version of R by typing `version` into the console and RStudio by typing `RStudio.Version()` into the console and update them if needed.

-   Windows

    -   Install and load installr:

        -   `install.packages("installr");library(installr)`\

    -   Copy/Update the existing packages to the new R installation:

        -   `updateR()`

    -   Open RStudio, go to Help \> Check for Updates to install newer
        version of RStudio (if available).

-   Mac

    -   Go to <https://cloud.r-project.org/bin/macosx/>.\
    -   Download the latest release (R-4.4.1.pkg) and finish the
        installation.
    -   Open RStudio, go to Help \> Check for Updates to install newer
        version of RStudio (if available).
    -   To update packages, go to Tools \> Check for Package Updates. If
        updates are available, select All, and click Install Updates.

#### 2. Required packages          

    -   `rmarkdown`
    -   `rstac`
    -   `terra`
    -   `imager`
    -   `leaflet`  
    -   `dygraphs`
    -   `xts`  
    -   `lubridate`  
    -   `earthdatalogin`
  
Run the cell below to identify any missing packages to install, and then load 
all of the required packages.

```{r, warning = FALSE, message = FALSE}
packages <- c('rmarkdown','earthdatalogin', 'rstac','imager','lubridate','xts','dygraphs','leaflet','terra')              

# Identify missing (not installed) packages
new.packages = packages[!(packages %in% installed.packages()[,"Package"])]

# Install new (not installed) packages
if(length(new.packages)) install.packages(new.packages, repos='http://cran.rstudio.com/', dependencies = TRUE) else print('All required packages are installed.')
```

# 1. Getting Started

## 1a. Load Required Libraries

Next load all packages using `library()` function.

```{r, warning= FALSE, message=FALSE}
invisible(lapply(packages, library, character.only = TRUE))
```

## 1b. Set Up the Working Directory

Create an output directory for the results.

```{r}
# Create an output directory if it doesn't exist
outDir <- file.path("data", "R_Output", fsep="/")
suppressWarnings(dir.create(outDir)) 
```

## 1c. Set Up the Authentication. 

The `earthdatalogin` sets up authentication using environmental variables. 

```{r}
# Authentication is not used needed immediately, but it is good to set up in the beginning.
earthdatalogin::edl_netrc()
```

# 2. CMR-STAC API: Searching for Items

We will use the CMR-STAC API search endpoint to query the LPCLOUD catalog for
HLS data by ROI and time period of interest. In order to retrieve STAC Items 
that match your criteria, you need to define your query parameters, which we 
will walk through below.

To learn how to navigate through the structure of a CMR-STAC Catalog and
define Search parameters, see [Getting Started with NASA's CMR-STAC
API](https://git.earthdata.nasa.gov/projects/LPDUR/repos/data-discovery---cmr-stac-api/browse).
Information on the specifications for adding parameters to a search query can
be found [here](https://github.com/radiantearth/stac-api-spec/tree/master/item-search#query-parameters-and-fields).

For this tutorial, we will use the [rstac](https://brazil-data-cube.github.io/rstac/index.html)
library to interact with the CMR-STAC API. 

Define the CMR-STAC endpoint, which we will connect to for our search.
```{r}
s = stac("https://cmr.earthdata.nasa.gov/stac/LPCLOUD/")
```

## 2a. Collection Query

We will need to assign the lists one or more Collection IDs we want to include
in the search query to a variable. Only STAC Items from the provided 
Collections will be searched. Here, we are interested in both HLS **L30** (Landsata
8 and 9) and **S30** (Sentinel-2A and B) collections. To learn how to access 
Collection IDs in CMR-STAC visit [Getting Started with NASA's CMR-STAC API](https://git.earthdata.nasa.gov/projects/LPDUR/repos/data-discovery---cmr-stac-api/browse).

```{r}
HLS_col <- list("HLSS30_2.0", "HLSL30_2.0")
```

## 2b. Spatial Query Search Parameter

We can assign our spatial region of interest by loading a GeoJSON file using the 
`terra` package. An example GeoJSON file is supplied in the 'Data' directory of 
the 'hls_tutorial_r' repository named `FieldBoundary.geojson`. Read the file in 
and use it to perform the spatial subset.  

```{r, results= "hide"}
roi <- terra::vect("data/Field_Boundary.geojson")
```

We will use the `leaflet` package to plot the agricultural fields boundary on 
top of a ESRI world imagery basemap.   

```{r}
leaflet() %>% 
  addPolygons(data = roi, fill = FALSE) %>% 
  addProviderTiles(providers$Esri.WorldImagery) %>% 
  addMiniMap(zoomLevelFixed = 5)
```

This map provides a visual representation of our study area. To perfor a 
spatial search using the CMR-STAC we need to define the "bounding box" using
the lower left and upper right coordinates of our area of interest. The `terra`
package offers a convenient solution to get this information. Below, we'll 
extract the extent of our input GeoJSON file and generate the necessary spatial
search parameters for precise data retrieval.

```{r}
roi_extent <- terra::ext(roi)
bbox <- c(roi_extent$xmin, roi_extent$ymin, roi_extent$xmax, roi_extent$ymax)
```

## 2c. Temporal Query Search Parameter

Next, we will need to define a temporal search query parameter. In our example, 
we will set the time period of interest for two months of August and September 2021. 
Note that the temporal ranges should be specified as a pair of date-time values 
separated by comma (,) or forward slash (/). Each date-time value must have the 
format of`YYYY-MM-DDTHH:MM:SSZ` (ISO 8601). Additional information on setting temporal 
searches can be found in the [NASA CMR Documentation](https://cmr.earthdata.nasa.gov/search/site/docs/search/api.html#temporal-range-searches).  

```{r}
roi_datetime <- '2021-08-01T00:00:00Z/2021-09-30T23:59:59Z'   # YYYY-MM-DDTHH:MM:SSZ/YYYY-MM-DDTHH:MM:SSZ
```


## 2d. Submit a Query with Our Search Criteria

Now we are all prepared to submit a request to the CMR-STAC endpoint! We will 
do this using the `rstac` package. The output will show all the available 
data matching our search criteria based on our datetime and bounding box.  

```{r}
items <- s %>%
    stac_search(collections = HLS_col,
                bbox = bbox,
                datetime = roi_datetime,
                limit = 100) %>%
    post_request()
print(items)
```

Each item returned is a `FeatureCollection`. We can explore an example of what 
is in a feature by selecting the first one.  

```{r}
View(items$features[[1]])
```

We can also view the assets of each item in the response, to see what data 
(bands) can be accessed.  

```{r}
assets <- items_assets(items)
print(assets)
```

Working with the first feature, we can extract the browse image URL and plot it.

```{r}
browse_image_url <- items$features[[1]]$assets$browse$href

browse <-load.image(browse_image_url)
plot(browse)
```

Our first view of NASA data from the cloud.

# 3. Accessing and Interacting with HLS Data

This section will demonstrate how to simplify interacting with our search results,
filter based on cloud cover, select assets (bands), stream the data to memory, 
subset by our region of interest, and create a time-series of NDVI.

To make this data easier to work with, we can start to place the information
into a spatial dataframe using the use the `items_as_sf()` function.

```{r}
sf_items <- items_as_sf(items)
View(sf_items)
```

More information can be added to the spatial dataframe by pulling information
from each feature using `sapply` to extract the properties of each feature, then
binding it to our spatial dataframe.

```{r}
# Retrieve Granule ID for each feature
granule_id <- sapply(items$features, function(feature) feature$id)
# Add as first column in sf_items
sf_items <- cbind(granule = granule_id, sf_items)
View(sf_items)
```

## 3a. Select Bands and Filter by Cloud Cover

We can also retrieve the asset URLs for the bands we are interested in. In this
case, our goal is to build an NDVI timeseries from the NIR and Red bands. We 
will also want the quality and Fmask layers. The band number is different for
NIR between the HLSL30 and HLSS30 products. Below you can find the band numbers 
we want to retrieve for each of the two products.   

-   HLSS30 (Sentinel 2):

    -   "narrow" NIR = B8A
    -   Red = B04
    -   Quality = Fmask

-   HLSL30 (Landsat):

    -   NIR = B05
    -   Red = B04
    -   Quality = Fmask

Additional information on HLS band allocations can be found [here](https://lpdaac.usgs.gov/documents/1118/HLS_User_Guide_V2.pdf).

Let's define a function to retrieve the URL for each of these bands and add
them as columns to the dataframe. Afterwards, run it.

```{r}
# Define a function to extract asset urls for selected bands
# # This also includes a check to ensure the correct bands are extracted
# # # depending on the collection (HLSL30 or HLSS30)
extract_asset_urls <- function(feature) {
    collection_id <- feature$collection
    if (collection_id == "HLSS30_2.0") {
        bands = c('B8A','B04','Fmask')
    } else if (collection_id == "HLSL30_2.0") {
        bands = c('B05','B04','Fmask')}
    sapply(bands, function(band) feature$assets[[band]]$href)
}
# Retrieve Asset URLs for each feature using our extract_asset_urls function and transpose them to columns
asset_urls <- t(sapply(items$features, extract_asset_urls))
View(asset_urls)
```

Notice that our function has returned a table of each asset, but that the name 
for the NIR is B8A. This is because Sentinel was the first item in the search 
results. We'll want to rename these for clarity as we add them to the spatial 
dataframe.

```{r}
colnames(asset_urls) <- c('nir', 'red', 'fmask')
sf_items <- cbind(sf_items, asset_urls)
View(sf_items)
```

The 'eo:cloud_cover' column contains a percentage cloud-cover of a scene, which we can use
to filter our results. We'll filter out any scenes with more than 30% cloud cover.

```{r}
# Filter based on cloud cover
sf_items <- sf_items[sf_items$eo.cloud_cover < 30,]
# Reset Row Indices
row.names(sf_items) <- NULL
View(sf_items)
```

## 3b. Subset HLS COGs Spatially and Stack HLS Data Layers

First, set up rgdal configurations to access the cloud assets that we are 
interested in. You can learn more about these configuration options [here](https://trac.osgeo.org/gdal/wiki/ConfigOptions).  

```{r, results= "hide"}
setGDALconfig("GDAL_HTTP_UNSAFESSL", value = "YES")
setGDALconfig("GDAL_HTTP_COOKIEFILE", value = ".rcookies")
setGDALconfig("GDAL_HTTP_COOKIEJAR", value = ".rcookies")
setGDALconfig("GDAL_DISABLE_READDIR_ON_OPEN", value = "EMPTY_DIR")
setGDALconfig("CPL_VSIL_CURL_ALLOWED_EXTENSIONS", value = "TIF")
```

Next, we define a function to open an HLS COG file from Earthdata Cloud and 
crop it to our region of interest. During the cropping phase we want to be sure
to reproject our region of interest to match the CRS of the HLS data.  

**One important thing to note about HLS data is that some scenes have the scale 
factor for bands located in the wrong section of the metadata. This means that 
the scale factor is not applied to the data when it is read by `terra::rast`. 
To correct this we will check the metadata for the scale factor and apply it if
necessary.**  

```{r}
# This function reads an HLS scene from a URL, applies the scale factor if necessary, and optionally crops and
# masks the scene based on a polygon. It requries the above GDAL configurations and a .netrc file. A .netrc
# can be created by running `earthdatalogin::edl_netrc()`.
open_hls <- function(url, roi = NULL) {
  # Add VSICURL prefix
  url <- paste0('/vsicurl/', url)
  # Retrieve metadata
  meta <- describe(url)
  # Check if dataset is Quality Layer (Fmask) - no scaling this asset (int8 datatype)
  is_fmask <- any(grep("Fmask", meta))
  # Check if Scale is present in band metadata
  will_autoscale <- any(grep("Scale:", meta))
  # Read the raster
  r <- rast(url)
  # Apply Scale Factor if necessary
  if (!will_autoscale && !is_fmask){
    print(paste("No scale factor found in band metadata. Applying scale factor of 0.0001 to", basename(url)))
    r <- r * 0.0001
  }
  # Crop if roi specified
  if (!is.null(roi)){
    # Reproject roi to match crs of r
    roi_reproj <- project(roi, crs(r))
    r <- mask(crop(r, roi_reproj), roi_reproj)
  }
return(r)
}
```

Let's test this function on the first red band in our spatial dataframe and 
plot the results.

```{r}
# Test opening and crop
red <- open_hls(sf_items$red[1], roi)
plot(red)
```

We can apply this function to all of the red, nir, and fmask scenes in our 
spatial dataframe, using `lapply` to place the results in a list. This cell may
take some time depending on internet speed, as we're loading 72 scenes from the
cloud and cropping them. 

```{r}
red_stack <- lapply(sf_items$red, open_hls, roi = roi)
nir_stack <- lapply(sf_items$nir, open_hls, roi = roi)
fmask_stack <- lapply(sf_items$fmask, open_hls, roi = roi)
```

------------------------------------------------------------------------

# 4. Processing HLS Data

Now we can start asking our science questions. First we define the NDVI 
function and then execute it on the data loaded into memory. After that, we can
perform quality filtering to screen out any poor-quality observations.

## 4a. Calculate NDVI

Create a function to calculate NDVI.

```{r}
calculate_NDVI <- function(nir, red){
  ndvi <- (nir-red)/(nir+red)
  return(ndvi)
}
```

Now we can calculate NDVI from the list of red and nir rasters.  

**Note that we get a warning about the CRS. This is because HLS-Landsat and 
HLS-Sentinel have slightly different WKT strings that define their projections. 
This is a known issue with HLS data. Although we get a warning here, the CRS is
the same.**

```{r}
# Calculate NDVI For All of our Scenes
ndvi_stack <- mapply(calculate_NDVI, nir_stack, red_stack, SIMPLIFY = FALSE)
# Rename the Scenes in our List
names(ndvi_stack) <- sf_items$datetime
# Create a single Rast Object from our list
ndvi_stacks <- terra::rast(ndvi_stack)
```

Now we plot! Let's start with the first item in NDVI time series.  

```{r, warning=FALSE, message=FALSE}
# Create a color palette 
pal <- colorNumeric(terrain.colors(n = 100), c(0,1) ,na.color = "transparent", reverse = TRUE)

leaflet() %>% 
    addProviderTiles(providers$Esri.WorldImagery) %>%
    addRasterImage(ndvi_stacks[[1]], color = pal, opacity = 1) %>%
    addPolygons(data = roi, fill = FALSE) %>%
    addMiniMap(zoomLevelFixed = 5) %>%
    leaflet::addLegend(pal = pal, values = c(0,1), title = "NDVI")

```

If you get an from this chunk, please update to the latest version of `leaflet`.  

------------------------------------------------------------------------

## 4b. Quality Filtering

For HLS v2.0 products, all quality information is included in the Fmask layer. 
This layer includes values corresponding to combinations of bits that represent
different quality descriptions at each location within the scene.  

Bits are ordered `76543210` and correspond to the following quality descriptions:  

|Bit Number|Mask Name|Bit Value|Description|
|----------|---------|---------|-----------|
|   7-6    | Aerosol |  11     |   High    |
|          | Level   |  10     |   Medium  |
|          |         |  01     |   Low     |
|          |         |  00     |   Clear   |
|----------|---------|---------|-----------|
|    5     | Water   |  1      |    Yes    |
|          |         |  0      |    No     |
|----------|---------|---------|-----------|
|    4     |  Snow/  |  1      |    Yes    |
|          |  Ice    |  0      |    No     |
|----------|---------|---------|-----------|
|    3     | Cloud   |  1      |    Yes    |
|          | Shadow  |  0      |    No     |
|----------|---------|---------|-----------|
|    2     | Cloud/  |  1      |    Yes    |
|          | Shadow  |  0      |    No     |
|          | Adjacent|         |           |
|----------|---------|---------|-----------|
|    1     | Cloud   |  1      |    Yes    |
|          |         |  0      |    No     |
|----------|---------|---------|-----------|
|    0     | Cirrus  |Reserved |    NA     |
|----------|---------|---------|-----------|

Open an fmask layer and plot it to see examples of the quality values.  

```{r}
# Test fmask
fmask <- fmask_stack[[23]]
plot(fmask)
```

We can select select the bits we want to mask from our analysis. In this case, 
we will mask pixels identified as clouds, cloud/shadow adjacent, cloud shadow, 
snow/ice, and water. We will ignore the aerosol levels. Create a list of those 
bit numbers:  

```{r}
selected_bit_nums <- c(1,2,3,4,5)
```

Now we can build a function to create a binary mask layer from the Fmask 
layers. This function does a couple things:  

1. Read the Fmask layer
2. Creates and empty mask Layer
3. Applies a bitwise AND operation to determine if the fmask value has the 
selected bit numbers number set when converted to binary.
4. If so, the mask value is set to 1, then the mask is updated using a bitwise
OR operation to update the layer for each selected bit as we loop through the
selected bit numbers.
5. Lastly, it returns the new mask layer we can use to filter our data.  

```{r}
# Filter based on quality
build_mask <- function(fmask, selected_bit_nums){
  # Create a mask of all zeros
  mask <- rast(fmask, vals=0)
  for (b in selected_bit_nums){
    # Apply Bitwise AND to fmask values and selected bit numbers
    mask_temp <- app(fmask, function(x) bitwAnd(x, bitwShiftL(1,b)) >0)
    # Update Mask to maintain only 1 layer with bitwise OR
    mask <- mask | mask_temp
  }
  return(mask)
}
```

To see this in action, we can apply the function to an Fmask layer and plot the 
results.  

```{r}
qmask <- build_mask(fmask[[1]], selected_bit_nums)
plot(qmask)
```

Like we did before, we can apply this function to our Fmask stack to create a 
list of masks.  

```{r}
# Create List of Masks
qmask_stack <- lapply(fmask_stack, build_mask, selected_bit_nums=selected_bit_nums)
```

After creating a list of masks, we can apply them to our NDVI stack to filter 
out the poor-quality pixels for each scene in the stack, making sure to update
the values masked values to `NA`.  

```{r}
# Apply Mask to NDVI using NA Values
ndvi_masked <- mapply(function(x, y) {
  mask(x, y, maskvalue = TRUE, updatevalue = NA)
}, ndvi_stack, qmask_stack, SIMPLIFY = FALSE)
```

Right now ndvi_masked is a list of `SpatRaster` objects. We can stack it using 
the `rast` function to provide the expected format for the next steps.  

```{r}
ndvi_masked <- rast(ndvi_masked)
```

Let's visualize a single layer of the masked NDVI to ensure our quality 
filtering worked.  

```{r}
plot(ndvi_masked[[23]])
```

We can see that some of the water present has been masked out of the NDVI layer,
which was part of our quality bit selection.  

## 4c.Visualize Stacked Time Series

Now we can plot multiple layers to create an interactive NDVI time series map 
with `leaflet`. Click on the dates on the left side to view the layer.  

```{r}
# Create a leaflet map and add the base layer
map <- leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addMiniMap(zoomLevelFixed = 5)

# Add each layer from our rasterstack
for (i in 1:nlyr(ndvi_masked)) {
  map <- map %>%
    addRasterImage(ndvi_masked[[i]], colors = pal, opacity = 1, group = names(ndvi_masked)[i])
}

# Add layer controls and legend
map <- map %>%
  addLayersControl(baseGroups = names(ndvi_masked),
                   options = layersControlOptions(collapsed = FALSE), position = 'topleft') %>%
  addLegend(pal = pal, values = c(0, 1), title = 'NDVI')
# Show map
map
```

Above, the time series show the changes in NDVI over the two months of September 
and August 2021. The NDVI values over these agricultural fields are high and 
stable and then they drastically decrease showing that they are cultivated 
before the send of September.  

## 4d. Calculating and Exporting Statistics

We can plot the time series NDVI values as boxplots showing their distribution 
for our farm fields.  

```{r}
# Add Date Only Column
sf_items$date <- sapply(sf_items$datetime, function(x) strsplit(x, "T")[[1]][1])
# Set Plot Margins and Font Sizes
par(mar = c(10, 6, 4, 2), cex.axis = 2, cex.lab = 2, cex.main = 2) # bottom, left, top, right
# Create Boxplot
terra::boxplot(ndvi_masked, col=c('olivedrab3'),  main='NDVI Time Series', ylab='NDVI',
                names = sf_items$date, las=2)
```

Next, calculate the statistics for each observation using built in statistics 
functions from `terra`.  

```{r}
ndvi_mean <- terra::global(ndvi_masked, 'mean', na.rm=TRUE)
ndvi_max <- terra::global(ndvi_masked, 'max', na.rm=TRUE)
ndvi_min <- terra::global(ndvi_masked, 'min', na.rm=TRUE)
ndvi_sd <- terra::global(ndvi_masked, 'sd', na.rm=TRUE)
```

With these stats, we can create an interactive plot using `dygraphs` library. 
We will leverage the `lubridate` to convert our dates to a more usable format, 
and the `xts` package to transform the dataframe to a format for `dygraph`.

```{r}
stats <- data.frame(
  NDVI_Max = ndvi_max,
  NDVI_Min = ndvi_min,
  NDVI_mean = ndvi_mean,
  NDVI_SD = ndvi_sd
)
stats$Date <- ymd_hms(sf_items$datetime) # convert string to date format (ISO 8601)
variables = xts(x=stats[,-5], order.by=stats$Date) # Choose the cols with the variables
dygraph(variables) %>%
  dyAxis("y",label = "NDVI")
```

If you want to export these statistics, we can do so to a CSV file.  

```{r}
stats_name <- file.path(outDir, "HLS_NDVI_Statistics.csv")
write.csv(stats,stats_name)
```

# 5. Export Output to GeoTIFF

Lastly, if you want to capture the final output files locally on your machine, 
you can export them as GeoTIFFs.  

```{r}
for (i in 1:nlyr(ndvi_masked)){
  file_name <- paste0("HLS.", gsub("[:]",".", names(ndvi_masked[[i]])), ".NDVI.tif")
  output_name <- file.path(outDir, file_name)
  terra::writeRaster(ndvi_masked[[i]], output_name, overwrite = TRUE)
}
```

The raster stack object can also be written to the disk.  

```{r, warning=FALSE}
output_name <- file.path(outDir, "HLS_NDVI_stack.tif")
terra::writeRaster(ndvi_masked, filename=output_name, overwrite=TRUE)
```

And we're done! You have successfully analyzed data in the cloud, exporting 
just the information you needed for your area of interest rather than having to
download everything.  

------------------------------------------------------------------------

### Contact Information
  
  Contact: [LPDAAC\@usgs.gov](mailto:LPDAAC@usgs.gov){.email}  
  Voice: +1-866-573-3222  
  Organization: Land Processes Distributed Active Archive Center (LP DAAC)  
  Website: <https://lpdaac.usgs.gov/>                   
  Date last modified: 09-17-2024                
  
  Work performed under USGS contract G0121D0001 for LP DAAC^1^.
  ^1^ LP DAAC Work performed under NASA contract NNG14HH33I.
